ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-12 paper/paper.tex
circles and gmsc

The findings for the Moons dataset are broadly in line with the findings so far: for the JEM, ECCCo yields significantly more faithful and plausible counterfactuals than all other generators. For the MLP, faithfulness is maintained but not plausibility. For comparison, REVISE yields fairly plausible counterfactuals in both cases, but it does so at the cost of faithfulness. For the Circles data, ...

Moving on to our real-world datasets, the results are shown in Table~\ref{tab:results-real-world}. Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs. The case is very clear for MNIST: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and Joint Energy modelling. For the JEM Ensemble, ECCCo is essentially on par with REVISE and does significantly better than the baseline generator.

The results for GMSC ...

>>>>

The findings for the Moons dataset are broadly in line with the findings so far: for the JEM, ECCCo yields significantly more faithful and plausible counterfactuals than all other generators. For the MLP, faithfulness is maintained but counterfactuals are not plausible. By comparison, REVISE yields fairly plausible counterfactuals in both cases, but it does so at the cost of faithfulness. We also observe that the best results for ECCCo are achieved when using both penalties. Once again though, the generative component (EBM) has a stronger impact on the positive results for the JEM. a

For the Circles data, the most faithful counterfactuals are generated by ECCCo. While it appears that REVISE generates the most plausible counterfactuals in this case, we note that they are valid only half of the time (see Appendix~\ref{app:results} for a complete overview of all evaluation metrics). It turns out that in this case, the underlying VAE with default parameters has not adequately learned the data-generating process. Of course, it is possible to achieve better generative performance through hyperparameter tuning. But this example serves to illustrate that REVISE depends strictly on the quality of the surrogate model. Independent of the outcome for REVISE, however, the results do not seem to indicate that ECCCo significantly improves our plausibility metric for the Circles data.

\textit{NOTE: visual inspection of the results shows that ECCCo still does exactly what it's supposed to do. I think the outcome here demonstrates the limitations of the implausibility metric: computing average distances to randomly drawn samples from the target class does not really make perfect sense when samples are distributed as a circle. But this story may be too convoluted to add here?}

Moving on to our real-world datasets, the results are shown in Table~\ref{tab:results-real-world}. Once again the findings indicate that the plausibility of ECCCos is positively correlated with the capacity of the black-box model to distinguish plausible from implausible inputs. The case is very clear for MNIST: ECCCos are consistently more faithful than the corresponding counterfactuals produced by any of the benchmark generators and their plausibility gradually improves through ensembling and joint-energy modelling. For the JEM Ensemble, ECCCo is essentially on par with REVISE and does significantly better than the baseline generator. We also note that ECCCo is the only generator that consistently achieves full validity for all models (Appendix~\ref{app:results}). Interestingly, ECCCo also yields lower-cost outcomes than the baseline generator for the JEMs.

For the tabular credit dataset (GMSC) we have struggled to get good generative and discriminative performance for our JEMs. Consequently, it is not surprising to find that ECCCo never achieves state-of-the-art plausibility, although it does improve outcomes compared to the baseline (Wachter). With respect to faithfulness, ECCCo once again consistently outperforms all other counterfactual generators.
